---
title: "Day 2"
date: 2020-12-02
author:
  name: https://adventofcode.com/2020/day/2
  url: https://adventofcode.com/2020/day/2
---

## Setup

Import libraries:

```{r setup}

library(tidyverse)
library(unglue)

```

Read and parse text input from file:

```{r}

input <- read_lines("../input/day02.txt") |> 
  unglue_data("{min}-{max} {chr}: {pwd}", convert = TRUE)

```

## Part 1

Count the occurrences of the designated character in each password and determine whether it falls within the bounds. Count how many passwords are valid.

```{r}

input |> 
  mutate(count = str_count(pwd, chr)) |> 
  filter(between(count, min, max)) |> 
  nrow()

```

## Part 2

Extract the characters at the indicated indices of each password, and count how many passwords have exactly 1 one of those positions equal to the desired character:

```{r}

input |> 
  mutate(
    idx = map2(min, max, c),
    str = str_sub_all(pwd, start = idx, end = idx),
    is_valid = map2_int(str, chr, ~ sum(.x == .y)) == 1
  ) |> 
  filter(is_valid) |> 
  nrow()

```

