---
title: "Day 3"
date: 2020-12-03
author:
  name: https://adventofcode.com/2020/day/3
  url: https://adventofcode.com/2020/day/3
---

## Setup

Import libraries:

```{r setup}

library(tidyverse)

```

Read text input from file:

```{r}

input <- read_lines("../input/day03.txt")

```

Parse character strings and convert into a 1/0 matrix:
 
```{r}

mtx <- input |> 
  str_split("") |> 
  map(~ case_match(.x, "." ~ 0, "#" ~ 1)) |> 
  reduce(rbind) |> 
  unname()

```

## Part 1

Define a function that converts a slope to indices:

```{r}

slope_to_idx <- function(mtx, right, down) {
  rows <- seq.int(1, nrow(mtx), by = down)
  cols <- ((1:length(rows) - 1) * right) %% ncol(mtx) + 1
  
  array(c(rows, cols), dim = c(length(rows), 2))
}

```

Define a function to count the encountered trees, given a slope:

```{r}

count_trees <- function(mtx, right, down) {
  sum(mtx[slope_to_idx(mtx, right, down)])
}

```

Count the encountered trees for a slope of right 3, down 1:

```{r}

count_trees(mtx, 3, 1)

```

## Part 2

Compare the encountered trees when using the following slopes and take their product:

```
Right 1, down 1
Right 3, down 1
Right 5, down 1
Right 7, down 1
Right 1, down 2
```

```{r}

right <- c(1, 3, 5, 7, 1)
down  <- c(1, 1, 1, 1, 2)

map2_dbl(right, down, \(right, down) count_trees(mtx, right, down)) |> 
  prod()

```

