---
title: "Day 5"
date: 2021-12-05
author:
  name: https://adventofcode.com/2021/day/5
  url: https://adventofcode.com/2021/day/5
---

## Setup

Import libraries:

```{r setup}

library(tidyverse)
library(unglue)

```

Read input from file:

```{r}

input <- read_lines("../input/day05.txt")

```

Convert plain-text input to a dataframe of numeric values:

```{r}

input_df <- input |> 
  unglue_data("{x1=\\d+},{y1=\\d+} -> {x2=\\d+},{y2=\\d+}", convert = TRUE) |> 
  mutate(id = row_number(), .before = everything())

```


## Part 1

Flag each line segment as horizontal, vertical or diagonal. Then, compute the list of all integer points that fall along the line.

```{r}

lines <- input_df |> 
  mutate(
    type = case_when(
      x1 == x2 ~ "v",
      y1 == y2 ~ "h",
      .default = "d"
    ),
    x = map2(x1, x2, ~ seq(..1, ..2)),
    y = map2(y1, y2, ~ seq(..1, ..2))
  ) |> 
  unnest(cols = c(x, y))

```

Compute the number of points that are covered by at least 2 overlapping horizontal or vertical lines (diagonals excluded):

```{r}

lines |> 
  filter(type %in% c("v", "h")) |> 
  summarize(n = n(), .by = c(x, y)) |> 
  filter(n > 1) |> 
  nrow()

```

## Part 2

Repeat the count of overlapping points, now allowing diagonals:

```{r}

lines |> 
  summarize(n = n(), .by = c(x, y)) |> 
  filter(n > 1) |> 
  nrow()

```



