---
title: "Day 7"
date: 2021-12-07
author:
  name: https://adventofcode.com/2021/day/07
  url: https://adventofcode.com/2021/day/07
---

## Setup

Import libraries:

```{r setup}

library(tidyverse)

```

Read input from file:

```{r}

input <- scan("../input/day07.txt", sep = ",", quiet = TRUE)

```

## Part 1

We define the cost function as the sum of the absolute differences between our given point and all crabs:

```{r}

cost <- \(x) sum(abs(x - input))

```

We define the set of candidates for the optimal alignment location:

```{r}

candidates <- min(input):max(input)

```

The total fuel cost is computed by minimizing the cost function over all candidates:

```{r}

candidates |> 
  map_dbl(cost) |> 
  min()

```

## Part 2

Instead of the basic absolute difference function, our new distance function for each crab $c_i$ is given by the binomial coefficient:

$$
1 + 2 + \cdots + |x - c_i| = \sum_{k = 1}^{|x - c_i|}k  = \binom{|x - c_i| + 1}{2}
$$

```{r}

cost <- \(x) sum(choose(abs(x - input) + 1, 2))

```

Repeating the computation with our new cost function, we get:

```{r}

candidates |> 
  map_dbl(cost) |> 
  min()

```
