---
title: "Day 25"
date: 2021-12-25
author:
  name: https://adventofcode.com/2021/day/25
  url: https://adventofcode.com/2021/day/25
---

## Setup

Import libraries:

```{r setup}

library(tidyverse)

```

Read and parse text input from file:

```{r}

input <- read_lines("../input/day25.txt")

```

## Part 1

Convert input to a matrix. Represent N/S-facing values as imaginary numbers and W/E-facing values as real numbers.

```{r}

mtx <- input |> 
  str_split("") |> 
  reduce(rbind) |> 
  unname()

h <- nrow(mtx)
w <- ncol(mtx)

```

Define functions to check if the value ahead of the current one to the south or east are blocked:

```{r}

is_blocked_s <- \(x) unname(rbind(tail(x, -1), head(x, 1))) != "."
is_blocked_e <- \(x) t(is_blocked_s(t(x)))

```

Define functions to shift the matrix values to the south or east

```{r}

move_s <- function(mtx) {
  blocked <- is_blocked_s(mtx)
  active  <- mtx == "v"
  
  old_pos <- which(active & !blocked, arr.ind = TRUE)
  
  new_pos <- old_pos
  new_pos[, "row"] <- (new_pos[, "row"] %% h) + 1
  
  mtx[new_pos] <- "v"
  mtx[old_pos] <- "."
  mtx
}

move_e <- function(mtx) {
  blocked <- is_blocked_e(mtx)
  active  <- mtx == ">"
  
  old_pos <- which(active & !blocked, arr.ind = TRUE)
  
  new_pos <- old_pos
  new_pos[, "col"] <- (new_pos[, "col"] %% w) + 1
  
  mtx[new_pos] <- ">"
  mtx[old_pos] <- "."
  mtx
}

move <- \(x) move_s(move_e(x))

```

Loop until we don't see any difference between the previous and current steps:

```{r}

n_steps <- function(mtx) {
  n <- 1
  prv <- mtx
  
  repeat {
    cur <- move(prv)
    
    if (all(cur == prv))
      return(n)
    
    prv <- cur
    n <- n + 1
  }
}

n_steps(mtx)

```


