---
title: "Day 17"
date: 2022-12-17
author:
  name: https://adventofcode.com/2022/day/17
  url: https://adventofcode.com/2022/day/17
---

## Setup

```{r setup}

# Libraries
library(tidyverse)

# Read input from file
input <- read_lines("../input/day17.txt", skip_empty_rows = TRUE) |> 
  str_split_1("") 

```

## Part 1

Representing obstacles as 1s and empty space as 0s, represent the shapes of the falling blocks, the floor, and the walls as bitwise integers:

```{r}

block1 <- strtoi(c("000111100"                                    ), base = 2)
block2 <- strtoi(c("000010000","000111000","000010000"            ), base = 2)
block3 <- strtoi(c("000111000","000001000","000001000"            ), base = 2)
block4 <- strtoi(c("000100000","000100000","000100000","000100000"), base = 2)
block5 <- strtoi(c("000110000","000110000"                        ), base = 2)

walls  <- strtoi("100000001", base = 2)
floor  <- strtoi("111111111", base = 2)

blocks <- list(block1, block2, block3, block4, block5)

```

Define functions that give properties of the tower:

```{r}

# Print tower to terminal (for debugging)
print_tower <- function(tower) {
  tower |>
  imap_chr(\(row, name) {
    row |> 
      intToBits() |> 
      rev() |> 
      tail(9) |> 
      as.integer() |> 
      case_match(0 ~ "·", 1 ~ "#") |> 
      modify_at(.at = c(1L, 9L), .f = ~ "|") |> 
      str_c(collapse = "") |> 
      str_c(name, sep = " ")
  }) |> 
  modify_at(.at = 1L, .f = ~ if_else(.x == "|#######| 0", "+-------+ 0", .x)) |> 
  rev() |> 
  cat(sep = "\n")
}

tower_height <- function(tower) {
  idx <- max(which(tower != walls))
  tower[idx] |>
    names() |> 
    as.double()
}

tower_base <- function(tower) {
  idx <- max(which(accumulate(tower, bitwOr, .dir = "backward") == floor))
  tower[idx] |>
    names() |> 
    as.double()
}

trim_tower <- function(tower) {
  base <- tower_base(tower)
  top  <- tower_height(tower)
  tower[as.character(base:top)]
}

```

Define functions that move blocks and check if the move is valid:

```{r}

shift_block <- function(block, dir) {
  f <- switch(dir, 
    "<" = bitwShiftL, 
    ">" = bitwShiftR
  )
  f(block, 1)
}

is_collision <- function(block, tower_slice) {
  any(bitwAnd(block, tower_slice) > 0)
}

# Try to move the block L/R if the move is valid, or return the old one if not
try_shift_block <- function(block, dir, tower_slice) {
  new <- shift_block(block, dir)
  if (is_collision(new, tower_slice))
    block
  else
    new
}

```

Define a function to drop blocks onto a tower:

```{r}

drop_blocks <- function(jets, n_blocks) {
  
  tower        <- floor
  names(tower) <- 1:length(tower) - 1
  
  n_jets <- length(jets)
  time   <- 0
  
  # Cycle through the list of blocks and drop them in order
  for (i in 1:n_blocks) {
    
    block_idx <- (i - 1) %% length(blocks) + 1
    block     <- blocks[[block_idx]]
    
    # Initialize the vertical location of the block
    block_loc <- 1:length(block) + 3 + tower_height(tower)
    
    # Add empty wall space to the top of the tower
    add_walls <- rep(walls, length(block) + 3) |> 
      set_names(c(min(block_loc) - 3:1, block_loc))
    tower <- c(tower, add_walls)
    
    # Drop block until it comes to rest
    repeat {
      jet_idx <- time %% n_jets + 1
      
      # Apply jet blast & increment time
      block <- try_shift_block(block, jets[jet_idx], tower[as.character(block_loc)])
      time <- time + 1
      
      # Check if block has come to rest; if so, add block to tower
      if (is_collision(block, tower[as.character(block_loc - 1)])) {
        tower[as.character(block_loc)] <- bitwOr(tower[as.character(block_loc)], block)
        tower <- trim_tower(tower)
        break
        
      } 
      # Otherwise drop block one unit and repeat the block jet seq
      else {
        block_loc <- block_loc - 1
      }
    }
  }
  
  tower
}

```

Run on puzzle input:

```{r}

input |> 
  drop_blocks(2022) |> 
  tower_height()

```

## Part 2

Modify the drop_blocks function to loop until a cycle is found and return cycle info:

```{r}

find_cycle <- function(jets) {
  
  tower        <- floor
  names(tower) <- 1:length(tower) - 1
  
  n_jets <- length(jets)
  time   <- 0
  states <- tibble(
    n_blocks  = numeric(0),
    tower_idx = list(),
    tower_val = list(), 
    block_idx = numeric(0), 
    jet_idx   = numeric(0)
  )
  
  i <- 1
  # Cycle through the list of blocks and drop them in order
  repeat {
    
    block_idx <- (i - 1) %% length(blocks) + 1
    block     <- blocks[[block_idx]]
    
    # Initialize the vertical location of the block
    block_loc <- 1:length(block) + 3 + tower_height(tower)
    
    # Add empty wall space to the top of the tower
    add_walls <- rep(walls, length(block) + 3) |> 
      set_names(c(min(block_loc) - 3:1, block_loc))
    tower <- c(tower, add_walls)
    
    # Drop block until it comes to rest
    repeat {
      jet_idx <- time %% n_jets + 1
      
      # Apply jet blast & increment time
      block <- try_shift_block(block, jets[jet_idx], tower[as.character(block_loc)])
      time <- time + 1
      
      # Check if block has come to rest; if so, add block to tower
      if (is_collision(block, tower[as.character(block_loc - 1)])) {
        tower[as.character(block_loc)] <- bitwOr(tower[as.character(block_loc)], block)
        tower <- trim_tower(tower)
        
        # Add block and jet index to the states list
        states <- states |> 
          add_row(
            n_blocks  = i,
            tower_idx = list(names(tower)),
            tower_val = list(unname(tower)),
            block_idx = block_idx,
            jet_idx   = jet_idx
          )
        # print(tower)
        break
        
      } 
      # Otherwise drop block one unit and repeat the block jet seq
      else {
        block_loc <- block_loc - 1
      }
    }
    
    i <- i + 1
    
    # After each block is dropped, check if a cycle has been found and return it
    dupes <- states |> 
      filter(n_distinct(tower_val) != n(), .by = c(block_idx, jet_idx))
    
    if (nrow(dupes) > 0) {
      
      cycle_length <- dupes |>
        pull(n_blocks) |>
        reduce(`-`) |>
        abs()

      cycle_start <- dupes |>
        pull(n_blocks) |>
        min()
      
      cycle_height <- dupes |>
        pull(tower_idx) |> 
        map(as.numeric) |> 
        reduce(`-`) |> 
        unique() |> 
        abs()

      return(list(length = cycle_length, start = cycle_start, height = cycle_height))
      return(dupes)
    }
  }
}

```

Get the cycle of the puzzle input:

```{r}

cycle <- find_cycle(input)

```

Using the cycle info from the output, compute the majority of the height using the cycle, then and add the height of the remaineder:

```{r}

n_cycles <- floor((1000000000000 - cycle$start) / cycle$length)
n_blocks <- (1000000000000 - cycle$start) %% cycle$length + cycle$start

((n_cycles * cycle$height) + tower_height(drop_blocks(input, n_blocks))) |> 
  format(scientific = FALSE)

```

