---
title: "Day 4"
date: 2024-12-04
author:
  name: https://adventofcode.com/2024/day/4
  url: https://adventofcode.com/2024/day/4
---

## Setup

Import libraries:

```{r setup}

library(tidyverse)

```

Read text input from file:

```{r}

input <- read_lines("../input/day04.txt")

```

Convert plain text lines into a character matrix:

```{r}

mtx <- input |> 
  map(~ matrix(str_split_1(.x, ""), nrow = 1)) |> 
  do.call(what = rbind)

```

## Part 1

Extract rows and columns as separate text strings from the matrix:

```{r}

mtx_rows <- apply(mtx, 1, str_flatten)
mtx_cols <- apply(mtx, 2, str_flatten)

```

Extract diagonals from the matrix by padding each row with leading/trailing spaces, then extracting the new columns as strings:

```{r}

padding <- str_dup(" ", 1:nrow(mtx))

mtx_diag <- map(c(I, rev), \(f) {
  str_c(padding, f(mtx_rows), rev(padding)) |> 
    str_split("") |> 
    list_transpose() |> 
    map_chr(str_flatten) |> 
    str_remove_all("\\s")
}) |> 
  unlist()

```

Search for the string `XMAS` (or its reverse, `SAMX`) in each dimension of the matrix and count the total occurrences:

```{r}

map(c("XMAS", "SAMX"), ~ str_count(c(mtx_rows, mtx_cols, mtx_diag), .x)) |> 
  unlist() |>
  sum()

```

## Part 2

Define the four possible XMAS patterns in a 3x3 grid as a regex string:

```{r}

xmas1 <- "M.S.A.M.S"
xmas4 <- "S.M.A.S.M"
xmas2 <- "S.S.A.M.M"
xmas3 <- "M.M.A.S.S"
xmas_regex <- str_glue("{xmas1}|{xmas2}|{xmas3}|{xmas4}")

```

Extract every 3x3 submatrix in the input text block as a text string (concatenating row-wise):

```{r}

sub_mtx <- expand_grid(row = 3:nrow(mtx), col = 3:ncol(mtx)) |> 
  pmap(\(row, col) mtx[row - 0:2, col - 0:2]) |> 
  map_chr(str_flatten)

```

Count all text strings with a valid `XMAS` pattern:

```{r}

sum(str_detect(sub_mtx, xmas_regex))

```

