---
title: "Day 23"
date: 2024-12-23
author:
  name: https://adventofcode.com/2024/day/23
  url: https://adventofcode.com/2024/day/23
---

## Setup

Import libraries:

```{r setup}

library(tidyverse)
library(unglue)
library(igraph)

```

Read input from text file into a data frame:

```{r}

input <- read_lines("../input/day23.txt") |> 
  unglue_data("{v1}-{v2}")

```

## Part 1

Convert the lists of connections to an undirected graph:

```{r}

g <- input |> 
  pmap(function(v1, v2) c(v1, v2)) |> 
  unlist() |> 
  make_graph(directed = FALSE)

```

Find all sets of 3 connected vertices (cliques of size 3) and count the sets having some vertex starting with 't'

```{r}

cliques(g, min = 3, max = 3) |> 
  keep(
    ~ names(.x) |> 
      str_starts("t") |> 
      any()
  ) |> 
  length()

```

## Part 2

Find the largest interconnected set of vertices, then convert the list of vertices into the "password" by sorting alphabetically and concatenating with commas:

```{r}

g |> 
  largest_cliques() |> 
  unlist() |> 
  names() |> 
  sort() |> 
  str_flatten(",")

```


